Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1) \note{\scriptsize Authors, please add/check middle initials and affiliations}

### 2014-04-13 WSSSPE1_report/WSSSPE1_summary_report.tex
Fine tune section 4.
Small changes to first half of paper.

\begin{abstract} %\note{Lead: Sou-Cheng T. Choi}
This document summarizes the First Workshop on Sustainable Software for Science:
Practice and Experiences (WSSSPE1). The content is based on 
a set of collaborative notes taken during the workshop,
including discussion about the keynote presentations, three themed panels that
discussed papers accepted at the workshop, other discussion, cross-cutting
issues, use cases, and conclusions. It joins a nascent literature seeking to
understand what drives software work in science and how the reward systems of
science function, thereby shaping the types of software projects undertaken, including the
extent to which developers are motivated to build software for the long-term,
for the use of others, and whether to work collaboratively or separately.
Unique perspectives were captured about issues such as comprehensive documentation, 
development and deployment practices; software licenses; career paths
for developers; identifiers for software through a
mechanism such as a Digital Object Identifier or publication of a ``software
paper''; and evidence of software contribution and impact, recorded in online
systems such as source code repositories like GitHub. For truly sustainable
software, there should be no endpoint, as the software products continue to be
used and useful beyond a single institution, grant, and developer or
development team.

@@

Because progress in scientific research is dependent on the quality of and
accessibility to software at all levels, it is now critical to address many
challenges related to the development, deployment, and maintenance of reusable
software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists,
researchers, and students are able to learn and adopt software-related skills
and methodologies. Established researchers are already acquiring some of these
skills, and in particular a specialized class of software developers is
emerging in academic environments as an integral part of
successful research teams. The WSSSPE workshop was intended to provide a forum
for discussion of the challenges, including both positions and experiences. The
short papers and debates have been archived to provide a basis for continued
discussion, and the workshop has fed into the collaborative writing of this
document. Estimates of 90 to 150 participants were present
throughout the day. Additional papers based on extended versions of workshop
submissions are expected. The level of interest in the workshop has led the
organizers, working with some of the submitters and attendees, to plan two
additional workshops (at the 2014 SciPy and SC14 conferences), to obtain
funds from the National Science Foundation and the Gordon and Betty Moore
Foundation to support these workshops, and to turn the
workshop website into a community website that can be used as a focus for
further discussion and progress. Additionally, a minisymposium that aims to
further explore some of the key issues raised in WSSSPE is co-organized by a
WSSSPE1 participant at the 2014 Society for Industrial and Applied Mathematics
(SIAM) Annual Meeting on ``Reliable
Computational Science.''

The goal of this paper is to summarize the contents of the workshop papers
and discussion, and identify issues that future WSSSPE events will address.
Before the WSSSPE1 workshop took place, the organizers self-published
a paper~\cite{WSSSPE1-pre-report} to document the process of
organizing and advertising the workshop, collecting and reviewing the
papers, and putting together the agenda. Section~\ref{sec:process} is
based on that publication. The remainder of this paper is based on the
workshop itself, as documented in a set of collaborative notes taken
during the workshop~\cite{WSSSPE1-google-notes}, including discussion
about the keynote presentations~(\S\ref{sec:keynotes}), the three
panels~(\S\ref{sec:devel}-\ref{sec:community}), 
cross-cutting issues~(\S\ref{sec:cross-cutting}),
use cases~(\S\ref{sec:use-cases}), and conclusions~(\S\ref{sec:conclusions}).  
Additionally, the accepted workshop papers are listed in
Appendix~\ref{sec:papers}, and a partial list of workshop attendees is
in Appendix~\ref{sec:attendees}.

@@

The organizers decided to list 54 of the papers (see Appendix~\ref{sec:papers} for a list of these papers) as significant
contributions to the workshop, a very high acceptance rate but one
that is reasonable, given the goal of broad participation and the fact
that the reports were already self-published. The papers were
grouped into three main categories, namely \emph{Developing and
Supporting Software}, \emph{Policy}, and \emph{Communities}. Each
subject area was associated with a panel and discussion at the
workshop.

The workshop itself consisted of two keynote presentations and the
three panels/discussion sessions. The panels were organized based on a
classification of the workshop submissions into three categories,
following the themes of the call for papers as modified by the areas
of the submissions. Three to four representatives from each submission
category were appointed as panelists, and assigned to read a subset of
the papers in that category and then discuss them in the panel.

@@

The first keynote~\cite{WSSSPE1-keynote1} was delivered by Philip
E. Bourne of University of California, San Diego.  Bourne is a basic
biomedical scientist but has also formed four software companies. He co-founded
PLOS Computational Biology~\cite{plos-web} and helped develop the RCSB
Protein Data Bank~\cite{pdb-web}.
He is working on automating three-dimensional visualizations of cell
contents and molecular structures, a problem that has not been solved
and when done, would serve as a key function of software in biomedical
sciences.

Bourne's keynote presentation was entitled ``A Recipe for Sustainable
Software,'' and developed based on his experiences.  He emphasized that
sustainability for software ``does not just mean more money from
Government'' (see also Section~\ref{sec:defining-sustainability}).  Other
factors to consider, he mentioned, include costs of production, ease
of maintenance, community involvement, and distribution channels.

@@

Bourne shared with the audience a few real challenges he
encountered. First, also an anecdote, he has looked into
reproducibility in computational biology, but has concluded that ``I
have proved I cannot reproduce research from my own
lab''~\cite{Veretnik}.

@@

He also had problems with selling research software: the technology
transfer office in his university wanted huge intellectual property
reach through, whereby they would get a share of profits from drugs
developed by pharmaceutical companies who use the software.  He was
aware this was unrealistic but the technology transfer office insisted
for a while. He wants to push a one-click approach for customers to
purchase university-written software.

He then presented arguments on directly valuing software as
a research output alongside papers, a common discussion within this field.
He mentioned an interesting exploration of involving software
engineers in the review process of scientific code~\cite{peer-review-code}.

On the notion of \emph{digital enterprise}, where information
technology (IT) underpins the whole of organizational activities, he
contended that universities are way behind the curve. In
particular, he highlighted the separation of research, teaching, and
administration into silos without a common IT framework as a blocker
to many useful organizational innovations: ``University 2.0 is yet to
happen.'' He spoke of a circumstance where someone had used an
algorithm developed for computational biology in marketing.
The role of an institution is important in this space. He argued that
funders can help train institutions, not just individuals, in this regard.

He concluded with a reference to his paper~\cite{bourne_ten} and
argued that computational scientists ``have a responsibility to
convince their institutions, reviewers, and communities that software is
scholarship, frequently
more valuable than a research article,'' a point with which all authors
strongly agree.

@@

The second keynote~\cite{WSSSPE1-keynote2} was delivered by Arfon
Smith of GitHub.

Smith's talk started with an example from data reduction in Astronomy,
where he needed to remove interfering effects from the device. He
built a ``bad pixel mask,'' and realized that while it was persistent,
there was no way or practice of sharing this with the
data. Consequently many researchers repeated the same calculations. He
estimated that 13 person-years were wasted in this redundant
calculation.

``Why didn't we do better?''  Smith asked of this practice. He argued
this was because we were taught to focus on immediate research
outcomes and not on continuously improving and building on tools for
research. He then asked, when we do know better, why we do not act any
different. He argued that it was due to incentives and their lack:
only the immediate products of research, not the software, are valued.
He referenced Victoria Stodden's talk at
OKCon~\cite{okcon-stodden-talk} which he said argued these points
well.

C. Titus Brown~\cite{ged-web}, a WSSSPE1 contributor and participant,
argued that with regard to reusable software, ``we should just start
doing it.'' In this regard Smith replied that documentation should be
``treated as a first class entity.''  He noted that the open source
community has excellent cultures of code reuse, for example,
RubyGems~\cite{rubygems-web}, PyPI~\cite{pypi-web}, and
CPAN~\cite{cpan-web}, where there is effectively low-friction
collaboration through the use of repositories. This has not happened
in highly numerical, compiled language scientific software.  An
exception he cited as a good example of scientific projects using
GitHub is the EMCEE Markov Chain Monte Carlo project~\cite{emcee-web}
by Dan Foreman-Mackey and contributors.

@@

The question of what ``sustainable'' means in the context of software was
raised (see \S\ref{sec:defining-sustainability}.) The resources at \url{http://oss-watch.ac.uk/resources/ssmm} discuss what
to look for when choosing software for procurement, or reuse in further software
development. Regardless of the license and development model that will be used,
the future of the project must be considered. Even if a particular piece of
software satisfies today's requirements, will it continue to do so in the
future? Important questions include whether the supplier will still be around in five
years time, will it still care for its customers needs, will it be responsive to
bug reports and feature requests? Should you tie your investment to a single
supplier using a proprietary product, or ensure the project uses an OSI-approved
license, and can outlive any single entity if the software is still important.

@@

The papers submitted to this panel, and several others, include lots of general
recommendations for processes, practices, tools, etc. One of the papers
suggested that a ``Software Sustainability Institute'' should be vested with
the authority to develop standardized quality processes, a central common
repository, central resources for services and consulting, a think tank of
sorts, and a software orphanage center. The idea of one common repository received
some resistance, with so many compelling alternatives available, e.g. Bitbucket or
GitHub. The point for centralization of communication/point of contact was seen as
reasonable, with the statement that ``vested with authority'' is perhaps too strong,
but ``providing tools if needed'' might be more appropriate.

@@

\subsection{Training Scientists to Develop and Support Software}

Some of the questions raised after the panel discussion are:

Rather than teaching developers about domain science, or domain scientists about
software development practices, why don't we teach both communities to
collaborate more effectively? Can this be done without teaching each side a
little of the other to enable communication, with a response that it really is
not two binary communities, but a spectrum with useful roles in the middle.
The question was raised if a developer can be effective without being part of the
domain community, with responses that this really depends on the specific
problem---translators and T-shaped people are important. Why aren't academic
communities taught how to evaluate cross-disciplinary work well?

What is the role for the growing field of team science? There
is overlap between the communities, with support for virtual organizations,
tool development, etc. How can we make time in an already crowded schedule to
introduce these topics to students? Should they be introduced through lab
classes as in ``real sciences''?

@@

There was debate that while some of the basics of computing may be fairly
similar tools commonly used for computing have changed quite
significantly. Reference was made to Perl, which was commonly used, giving way
to whole new languages, such as Python, for gluing things together and how this
induces many students into entirely rewriting the scaffolding, leaving the old
to rot and the experiments to become non-reproducible as the tools change.
Jason Riedy stated that he had been guilty of this in the past. There was
discussion of this tendency along with the enormous differences in the speed and
ease of sharing---having to ship tapes around (GCC, Emacs, etc) as opposed to
the immediate sharing of the latest development online, using revision control
systems like CVS, Subversion, Git, Mercurial, Bazaar, etc.

@@

The Policy panel was designated to discuss the group of workshop
contributions dealing with the wide range of software sustainability
aspects that relate to establishing, promoting, and implementing
policies. Six papers presented frameworks for defining, modeling, and
evaluating software sustainability, the basis of establishing
policies. Four papers advocated mechanisms for more properly assessing
the impact of scientific software, and for crediting and recognizing
work that promotes software sustainability, all of which are
instrumental in effectively promoting policies that aim to change
current practices. Four papers discuss facets of implementing software
sustainability, and models of implementation across different facets.

@@

How work on scientific software is recognized and rewarded strongly influences the motivation for particular kinds of work
on scientific software. A recurring theme of the panel discussion was
that software work in science is
inadequately visible within the reputation
system underlying science; in other words it often doesn't ``count''. In his paper for this workshop, Katz placed
software work along with other ``activities that facilitate science
but are not currently rewarded or
recognized''~\cite{Katz_WSSSPE}. Priem and Piwowar argued for the need
to ``support all researchers in presenting meaningful impact evidence
in tenure, promotion, and funding applications.''~\cite{Priem_WSSSPE}.
Kneply et al. argued that the visibility of software that supported a
piece of science ``can have detrimental effects on funding, future
library development, and even scientific
careers.''~\cite{Knepley_WSSSPE}.

@@

The literature below, suggested by attendees, addresses both non-scientific
software projects, as well as scientific projects.  The selections address both
descriptive and prescriptive approaches to communities.

Both \cite{howison_scientific_2011} and \cite{howison_incentives_2013} 
study how scientific software collectives self-organize and address issues of
incentive, long-term support, and development of infrastructure as well as new
features.  As noted elsewhere in this summary,
\cite{Turk:2013:SCH:2484762.2484782} shared prescriptions from two software
communities in astrophysics.

From the perspective of developing prescriptions for successful scientific
software development, both \cite{citeulike:11831265} and
\cite{1749-4699-6-1-015010} share experiences and suggestions for developing
sustainable practices.  \cite{citeulike:11831265} proposes "ten simple rules"
for developing open source scientific software, focusing on both the choices
made during development and the sustainability of practices in the long term.  
\cite{1749-4699-6-1-015010} describes the development and long-term growth of
the \texttt{deal.II} library, and how its place in its ecosystem of libraries,
applications and domains has shaped its development and community trajectory.

From more traditional open source development, resources were shared that
developed communities explicitly, such as 
\cite{citeulike:7888211} and \cite{Trapani:2011}, focusing on
large-scale projects such as the Ubuntu linux distribution and smaller-scale
volunteer-developed projects like such as ThinkUp, respectively.
The process of open source development, while less explicitly focused on
community building, outlined in \cite{citeulike:478633} was seen as a valuable
resource, particularly when combined with the management and personal
interaction techniques outlined in \cite{opac-b1134063}.  Growing diversity in
communities was directly addressed in \cite{Allsopp:2012}, where experiences
growing the diversity of technical conferences in open source were described.

@@

The papers describing education and training were focused primarily on how
these aspects of community development impact on the long-term sustainability
of software projects.  \cite{Girotto_WSSSPE} described the impact of the
mandate within the International Centre for Theoretical Physics (ICTP) to
foster resources and competences in software development and HPC, resulting in
the development of educational curricula directed in this area.  The paper
itself described the changes made in these curricula as a result of the current
changes in the HPC and scientific software landscape due to the advent of
scripting languages, new programming paradigms and new types of hardware such
as accelerator technologies.  The development of a workshop, with carefully
selected participants and an immersive approach to learning, was identified as
a major success for educating and developing new scientific software developers
from targeted domains.

\cite{Crawford_WSSSPE} addressed education and training within computational
chemistry frameworks, particularly as these frameworks attempt to address
next-generation computer hardware and software and as chemistry courses
emphasize lab work over computational education.  The authors identify this
lack of computational awareness and training as the primary challenge to future
advances in computational chemistry.  The authors propose a new institute for
computational chemistry, emphasizing collaboration (and a licensing structure,
such as LGPL or more permissive) and education of future generations of
chemistry researchers.

>>>>

\begin{abstract} %\note{Lead: Sou-Cheng T. Choi}
This document summarizes the First Workshop on Sustainable Software for Science:
Practice and Experiences (WSSSPE1). The content is based on the workshop
itself, as recorded in a set of collaborative notes taken during the workshop,
including discussion about the keynote presentations, three themed panels that
discussed papers accepted at the workshop`, other discussion, cross-cutting
issues, use cases, and conclusions. It joins a nascent literature seeking to
understand what drives software work in science and how the reward systems of
science work, thereby shaping the type of software work undertaken, including the
extent to which developers are motivated to build software for the long-term,
for the use of others, and whether to work collaboratively or separately.
Unique perspectives were captured about issues such as comprehensive documentation,
software deployment, and development practices; software licenses; career paths
for developers; identifiers for software through a
mechanism such as a Digital Object Identifier or publication of a ``software
paper''; and evidence of software contribution and impact, recorded in online
systems such as source code repositories like GitHub. For truly sustainable
software, there should be no endpoint, as the software products continue to be
used and useful beyond a single institution, grant, and developer or
development team.

@@

Because progress in scientific research is dependent on the quality of and
accessibility to software at all levels, it is now critical to address many
challenges related to the development, deployment, and maintenance of reusable
software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists,
researchers, and students are able to learn and adopt software-related skills
and methodologies. Established researchers are already acquiring some of these
skills, and in particular a specialized class of software developers is
emerging in academic environments who are an integral and embedded part of
successful research teams. The WSSSPE workshop  provided a forum
for discussion of the challenges around scientific software, including both positions and experiences. The
workshop short papers and debates have been archived to provide a basis for continued
discussion, and have fed into the collaborative writing of this
document. Additional papers based on extended versions of workshop
submissions have been developed as part of this special journal edition. The level of interest in the workshop which involved 
an estimated 90 to 150 participants during the day has led to further activities:
Two additional WSSSPE workshops are planned  (at the 2014 SciPy and SC14 conferences), 
funds to support the workshops have been obtained from 
the National Science Foundation and the Gordon and Betty Moore
Foundation, and the
original workshop website has been turned into a community website for
further discussion and progress. Additionally, a minisymposium  to
further explore some of the key issues raised in WSSSPE is co-organized by a
WSSSPE1 participant at the 2014 Society for Industrial and Applied Mathematics
(SIAM) Annual Meeting on ``Reliable
Computational Science.''

This paper summarizes the  workshop papers
and discussion, and identifies issues that future WSSSPE events will address.
Before the WSSSPE1 workshop took place, the organizers self-published
a paper~\cite{WSSSPE1-pre-report} to document the process of
organizing and advertising the workshop, collecting and reviewing the
papers, and putting together the agenda. Section~\ref{sec:process} is
based on that publication. The remainder of this paper is based on the
workshop itself, as documented in a set of collaborative notes taken
during the workshop~\cite{WSSSPE1-google-notes}, including discussion
about the keynote presentations (\S\ref{sec:keynotes}), the three
panels (\S\ref{sec:devel}-\ref{sec:community}), 
cross-cutting issues (\S\ref{sec:cross-cutting}),
use cases (\S\ref{sec:use-cases}), and conclusions
(\S\ref{sec:conclusions}).  Additionally, the accepted workshop papers are listed in
Appendix~\ref{sec:papers}, and a partial list of workshop attendees is
in Appendix~\ref{sec:attendees}.

@@

The organizers decided to list 54 of the papers (see Appendix~\ref{sec:papers} for a list of these papers) as significant
contributions to the workshop, a very high acceptance rate but one
that is reasonable, given the goal of broad participation and the fact
that the reports were already self-published. Based on the submissions, the papers were
grouped into three main categories, namely \emph{Developing and
Supporting Software}, \emph{Policy}, and \emph{Communities}. Each
category was associated with a panel and discussion at the
workshop.  Three to four representatives from each submission
category were appointed as panelists, and assigned to read a subset of
the papers in that category and then discuss them in the panel.
The workshop itself consisted of two keynote presentations followed by the 
three panel/discussion sessions.

@@

The first keynote~\cite{WSSSPE1-keynote1} was delivered by Philip
E. Bourne of University of California, San Diego.  Bourne is a 
biomedical scientist who has also formed four software companies. He co-founded
PLOS Computational Biology~\cite{plos-web} and helped develop the RCSB
Protein Data Bank~\cite{pdb-web}.
He is working on automating three-dimensional visualizations of cell
contents and molecular structures, a problem that has not been solved
and when done, would serve as a key function of software in biomedical
sciences.

Bourne's  presentation was based on his own software experiences.  He emphasized that
sustainability for software ``does not just mean more money from
Government'' (see also Section~\ref{sec:defining-sustainability}).  Other
factors to consider, he mentioned, include costs of production, ease
of maintenance, community involvement, and distribution channels.

@@

Bourne shared with the audience a few real challenges he
encountered. His first anecdote was that he has looked into
reproducibility in computational biology, but has concluded that ``I
have proved I cannot reproduce research from my own
lab''~\cite{Veretnik}.

@@

He also had problems with selling research software: the university technology
transfer office wanted huge and unrealistic intellectual property
reach through, whereby they would get a share of profits from drugs
developed by pharmaceutical companies who use the software. 
He advocated for  a one-click approach for customers to
purchase university-written software.

He then presented arguments on directly valuing software as
a research output alongside papers, a common discussion within this field.
He mentioned an  exploration of involving software
engineers in the review process of scientific code~\cite{peer-review-code}, and discussed 
how publishing software reviews could change attitudes.

On the notion of \emph{digital enterprise}, where information
technology (IT) underpins the whole of organizational activities, he
contended that universities are way behind the curve. In
particular, he highlighted the separation of research, teaching, and
administration into silos without a common IT framework as a blocker
to many useful organizational innovations: ``University 2.0 is yet to
happen.''  He argued that
funders such as NSF and NIH can help train institutions, not just individuals, in this regard.

He concluded by discussing his 2011 paper ``Ten Simple Rules for Getting Ahead as a Computational Biologist in Academia''~\cite{bourne_ten} and
argued that computational scientists ``have a responsibility to
convince their institutions, reviewers, and communities that software is
scholarship, frequently
more valuable than a research article''.

@@

The second keynote~\cite{WSSSPE1-keynote2} was delivered by Arfon
Smith of GitHub. Smith's talk started with an example from data reduction in Astronomy,
where he needed to remove interfering effects from the device (XXXX? which device). He
built a ``bad pixel mask,'' and realized that while it was persistent,
there was no way or practice of sharing this with the
data. Consequently many researchers repeated the same calculations. He
estimated that 13 person-years were wasted in this redundant
calculation. (XXXX whole paragraph not clear)

``Why didn't we do better?''  Smith asked of this practice. He argued
this was because we were taught to focus on immediate research
outcomes and not on continuously improving and building on tools for
research. He then asked, when we do know better, why we do not act any
different. He argued that it was due to the lack of incentives:
only the immediate products of research, not the software, are valued.
He referenced Victoria Stodden's talk at
OKCon~\cite{okcon-stodden-talk} which he said argued these points
well.

C. Titus Brown~\cite{ged-web}, a WSSSPE1 contributor,
argued that with regard to reusable software, ``we should just start
doing it.'' In this regard Smith replied that documentation should be
``treated as a first class entity.''  He noted that the open source
community has excellent cultures of code reuse, for example,
RubyGems~\cite{rubygems-web}, PyPI~\cite{pypi-web}, and
CPAN~\cite{cpan-web}, where there is effectively low-friction
collaboration through the use of repositories. This has not happened
in highly numerical, compiled language scientific software.  An
exception he cited as a good example of scientific projects using
GitHub is the EMCEE Markov Chain Monte Carlo project~\cite{emcee-web}
by Dan Foreman-Mackey and contributors.

@@

The question of what ``sustainable'' means in the context of software was
raised (see \S\ref{sec:defining-sustainability}.) The resources at OSS Watch~\cite{osswatch}   discuss what
to look for when choosing software for procurement, or reuse in further software
development. Regardless of the license and development model that will be used,
the future of the project must be considered. Even if a particular piece of
software satisfies today's requirements, will it continue to do so in the
future? Important questions include whether the supplier will still be around in five
years time, will it still care for its customers needs, will it be responsive to
bug reports and feature requests? Should you tie your investment to a single
supplier using a proprietary product, or ensure the project uses an OSI-approved
license, and can outlive any single entity if the software is still important.

@@

The papers submitted to this panel included many general
recommendations for processes, practices, tools, etc. One of the papers
suggested that a ``Software Sustainability Institute'' should be vested with
the authority to develop standardized quality processes, a central common
repository, central resources for services and consulting, a think tank of
sorts, and a software orphanage center. The idea of one common repository received
some resistance, with so many compelling alternatives available, e.g. Bitbucket or
GitHub. The point for centralization of communication/point of contact was seen as
reasonable, with the statement that ``vested with authority'' is perhaps too strong,
but ``providing tools if needed'' might be more appropriate.

@@

\subsection{Training Scientists to Develop and Support Software}
Part of the panel discussion focused around community structures and how academic communities are not taught how to evaluate cross-disciplinary work. 
One question raised was whether a software developer can be effective if they are not part
of the appropriate domain community, with responses that this depends on the 
specific problem and situation, and that "T-shaped" people who have both disciplinary depth as well as interdisciplinary and collaboration skills are important. 
The discussion focused on  whether we could teach software developers and domain scientists to 
collaborate together more effectively rather than trying to teach software developers
about domain science and domain scientists about software development practices. 
The end goal of this would be to have a single community with a spectrum of expertise across
domain science and software development rather than two separate communities.

The role of the growing field of team science with software development  was discussed. The science of team science
deals with understanding and improving collaborative and team-based scientific research, and such issues as virtual organizations, 
and tool development cut across software development communities. Further, how should these skills and best practices then be 
introduced to students?

@@

There was debate that while some of the basics of computing may be fairly
similar tools commonly used for computing have changed quite
significantly. Reference was made to Perl, which was commonly used, giving way
to whole new languages, such as Python, for gluing things together and how this
induces many students into entirely rewriting the scaffolding, leaving the old
to rot and the experiments to become non-reproducible as the tools change.
 There was
discussion of this tendency along with the enormous differences in the speed and
ease of sharing---having to ship tapes around (GCC, Emacs, etc) as opposed to
the immediate sharing of the latest development online, using revision control
systems like CVS, Subversion, Git, Mercurial, Bazaar, etc.

@@

The  panel on policy  discussed workshop
contributions dealing with the wide range of software sustainability
aspects that relate to establishing, promoting, and implementing
policies. Six papers presented frameworks for defining, modeling, and
evaluating software sustainability, the basis of establishing
policies. Four papers advocated mechanisms for more properly assessing
the impact of scientific software, and for crediting and recognizing
work that promotes software sustainability, all of which are
instrumental in effectively promoting policies that aim to change
current practices. Four papers discuss facets of implementing software
sustainability, and models of implementation across different facets.

@@

How work on scientific software is recognized and rewarded strongly influences the motivation for particular kinds of work
on scientific software. A recurring theme of the panel discussion was
that software work in science is
inadequately visible within the reputation
system underlying science; in other words it often doesn't ``count''. In his paper for this workshop, Katz placed
software work along with other ``activities that facilitate science
but are not currently rewarded or
recognized''~\cite{Katz_WSSSPE}. Priem and Piwowar argued for the need
to ``support all researchers in presenting meaningful impact evidence
in tenure, promotion, and funding applications.''~\cite{Priem_WSSSPE}.
Knepley et al. argued that the lack of visibility of software that supported a
piece of science ``can have detrimental effects on funding, future
library development, and even scientific
careers.''~\cite{Knepley_WSSSPE}.

@@

The literature below, suggested by attendees, addresses both non-scientific
software projects, as well as scientific projects.  The selections address both
descriptive and prescriptive approaches to communities. 
\begin{itemize}
\item \cite{howison_scientific_2011, howison_incentives_2013}
\item \cite{citeulike:11831265}
\item \cite{citeulike:7888211}
\item \cite{citeulike:478633}
\item \cite{opac-b1134063}
\item \cite{Trapani:2011}
\item \cite{Allsopp:2012}
\item \cite{1749-4699-6-1-015010}
\item \cite{Turk:2013:SCH:2484762.2484782}
\end{itemize}

@@

\todo{text about papers and discussion in this part of the workshop}
